---
title: "Assignment 5: Substance Use Treatment Mapping"
subtitle: "Faith Muller"
author: "Amber Camp"
date: "`r Sys.Date()`"
format: html
editor: visual
toc: true
---

## Overview

Due: **Friday 10/2, 11:59pm**

In this assignment, you will apply analytic and mapping skills to the 2024 TEDS-D dataset from SAMHSA.

Refer to the following files for reference, though some of what is asked of you here will require logic beyond what we discussed in class:

- week06-treatment-mapping-part1.qmd

- week06-treatment-mapping-part2.qmd

**Before you start:** keep the [codebook](https://www.samhsa.gov/data/system/files/media-puf-file/TEDS-D-2024-DS0001-info-codebook_v1_0.pdf) open in another tab. Every question below depends on knowing what the values in a column actually mean. Don't assume — look it up.

**Time estimate:** about three-four hours total. Question 1 should take you ten minutes. If you are stuck on one thing for more than twenty minutes, ask your classmates or get in touch with me. Being stuck is normal. Staying stuck quietly by yourself is not :)

## Load Packages and Data

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(sf) # "simple features" for spatial data
library(urbnmapr) # access to US state/country shapefiles for mapping
library(ggiraph) # makes ggplot charts interactive
options(scipen = 99) # turns off scientific notation
```

This is the cleaned dataset you saved at the end of Part 1.

```{r}
#| label: read-data

teds_d <- readRDS("data/teds_d_clean.rds")
```

## Question 1: Warm-up

Before mapping anything, get oriented in the data.

a)  Using `tabyl()`, show the breakdown of `reason` (reason for discharge). Which reason is most common, and what percentage of all discharges does it account for?

b)  Make one bar chart showing the number of discharges by `services_d` (type of treatment service). Give it an informative title and axis labels.

c)  In a sentence or two: what does the `reason` value "Transferred to another treatment program or facility" mean for our completion percentages? If a state transfers a lot of patients, would its completion rate look higher or lower than a state that transfers very few?

> **Full credit:** a working `tabyl()`, a readable labeled bar chart, and a sentence or two showing you thought about part (c). There is no trick in this question!

```{r}
#| label: q1
tabyl(teds_d$reason)

ggplot(data = teds_d, aes(x = services_d, color = reason)) +
         geom_bar() +
         labs(
           title = "The number of Discharges by type of treatment service",
           x     = "Treatment service",
           y     = "Number of Discharges") +
  theme(axis.text.x = element_text (angle = 45, hjust = 1, vjust = 1))
```

**Answer to (a): The most common reason is Treatment completed, it accounts for 47.8% of all discharges.**

**Answer to (b):** (produce a bar chart in the code chunk above)

**Answer to (c): It would look lower because transferred patients are counted as not completing, since they finish treatment somewhere else.**

## Question 2: Mapping no use at discharge

Make **one [*interactive*]{.underline} map** with `ggiraph` showing, for each state, the percentage of [completed treatment]{.underline} episodes that ended with [no use at discharge]{.underline}.

In other words: among the episodes where `reason` is "Treatment completed," what share had `freq1_d` equal to "no use"?

Work through this in bite-sized steps, like we did in Part 2. You will need to filter, group, summarize, build an index, join, and plot. Write one step at a time and look at the result before moving on.

### The denominator decision

About a third of the `freq1_d` values in this dataset are "unknown." That means you have to decide what your denominator is, and there is more than one defensible answer:

- **All** completed episodes, counting "unknown" in the denominator, or

- Only completed episodes where frequency of use was **actually recorded**.

Pick one. **Below your map, write one sentence saying which denominator you used and why.** Either choice can earn full credit. Missing description earns no points for this portion.

### Your map must:

- be interactive (rendered with `girafe()`)

- show a tooltip with the state name and the percentage, rounded to two decimal places

- have a title and a legend label that say what is being shown

- handle states with no data honestly (remember Friday in class)

> **Full credit:** one interactive map meeting the four requirements above, plus your denominator sentence. It does not have to be beautiful, only correct and readable.

```{r}
#| label: q2
#| warning: false

no_use_at_discharge <- teds_d |>
  filter(reason == "Treatment completed") |>
  group_by(stfips) |>
  summarize(
    total  = n(),
    no_use = sum(freq1_d == "no use")
  ) |>
  filter(no_use > 0) |>
  mutate(
    pct_no_use = no_use / total * 100,
    state_fips = sprintf("%02d", stfips)
  )

states_sf <- get_urbn_map("states", sf = TRUE)

no_use_map <- left_join(states_sf, no_use_at_discharge, by = "state_fips")

no_use_int <- ggplot(no_use_map) +
  geom_sf_interactive(
    aes(geometry = geometry,
        fill = pct_no_use,
        tooltip = paste("State:", state_name,
                        "<br>State FIPS:", stfips,
                        "<br>No use at discharge:", round(pct_no_use, 2), "%")),
    color = "black", linewidth = 0.25) +
  labs(fill = "% of Completed Episodes\nwith No Use at Discharge",
       title = "No Use at Discharge Among Completed Treatment Episodes",
       subtitle = "TEDS-D 2024 (SAMHSA). Gray states did not report data.") +
  scale_fill_viridis_c(option = "mako", direction = -1, na.value = "lightgrey") +
  coord_sf(datum = NA) +
  theme_minimal() +
  theme(panel.background = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        legend.text = element_text(size = 4),
        legend.title = element_text(size = 5))

girafe(ggobj = no_use_int)
```

**Denominator I used, and why:** I used all completed episodes, also including the ones with "unknown", so that every completed episode in a state is counted. I did this because I found it easier than doing it the other way.

Question 3: Service type and length of stay

Treatment outcomes can vary depending on the type of service a patient receives and how long they stay in treatment.

Investigate how **completion rates** differ across **service type** (`services_d`) and **length of stay** (`los`).

- Create **two** visualizations: at least one focused on service type, and at least one focused on length of stay. These do not need to be maps, and they do not need to be interactive. Pick whatever chart type makes the pattern easiest to see.

- Then write a short summary (3-ish sentences, full grammatical sentences) explaining the main patterns you observe.

**One warning:** look up `los` in the codebook *before* you calculate anything with it. Code that runs without an error can still give you a meaningless answer.

> **Full credit:** two labeled visualizations that actually show the comparison you say they show, plus a summary that describes a real pattern in your own output. You are not graded on finding the "right" pattern. You are graded on whether your summary matches your charts.

```{r}
#| label: q3-service

completion_by_service <- teds_d |>
  filter(services_d != "unknow") |>   # drops missing; matches your spelling from Monday
  group_by(services_d) |>
  summarize(
    total     = n(),
    completed = sum(reason == "Treatment completed", na.rm = TRUE)
  ) |>
  mutate(pct_completed = completed / total * 100)

completion_by_service

service_p <- ggplot(completion_by_service,
                    aes(x = fct_reorder(services_d, pct_completed),
                        y = pct_completed)) +
  geom_col_interactive(
    aes(tooltip = paste0(services_d,
                         "<br>Completed: ", round(pct_completed, 1), "%",
                         "<br>Discharges: ", scales::comma(total))),
    fill = "#2a6f97", width = 0.7) +
  geom_text(aes(label = paste0(round(pct_completed), "%")),
            hjust = -0.15, size = 3.2) +
  geom_hline(yintercept = 47.8, linetype = "dashed", color = "grey40") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.02)),
                     labels = function(x) paste0(x, "%")) +
  scale_x_discrete(labels = scales::label_wrap(30)) +
  labs(title    = "Completion Rate by Type of Treatment Service",
       subtitle = "TEDS-D 2024 (SAMHSA). Dashed line = national average (47.8%)",
       x = NULL,
       y = "% of discharges completed") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        plot.title         = element_text(face = "bold"))

girafe(ggobj = service_p, width_svg = 8, height_svg = 5)
```

```{r}
#| label: q3-los

completion_by_los <- teds_d |>
  mutate(los_group = cut(
    los,
    breaks = c(0, 1, 7, 30, 32, 34, 36, 37),
    labels = c("1 day", "2–7 days", "8–30 days", "31–60 days",
               "61–120 days", "121–365 days", "Over 1 year")
  )) |>
  filter(!is.na(los_group)) |>
  group_by(los_group) |>
  summarize(
    total     = n(),
    completed = sum(reason == "Treatment completed", na.rm = TRUE)
  ) |>
  mutate(pct_completed = completed / total * 100)

completion_by_los

los_p <- ggplot(completion_by_los, aes(x = los_group, y = pct_completed)) +
  geom_col_interactive(
    aes(tooltip = paste0(los_group,
                         "<br>Completed: ", round(pct_completed, 1), "%",
                         "<br>Discharges: ", scales::comma(total))),
    fill = "#2a6f97", width = 0.7) +
  geom_text(aes(label = paste0(round(pct_completed), "%")),
            vjust = -0.5, size = 3.2) +
  geom_hline(yintercept = 47.8, linetype = "dashed", color = "grey40") +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.02)),
                     labels = function(x) paste0(x, "%")) +
  labs(title    = "Completion Rate by Length of Stay",
       subtitle = "TEDS-D 2024 (SAMHSA). Dashed line = national average (47.8%)",
       x = "Length of stay",
       y = "% of discharges completed") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank(),
        plot.title         = element_text(face = "bold"))

girafe(ggobj = los_p, width_svg = 8, height_svg = 5)
```

**Summary:** I noticed that some types of treatments had much higher completion rates than others. Detox, 24-hour, hospital inpatient had the highest rate (77.3%), and Ambulatory, intensive outpatient had the lowest (29.9%). The completion percentage stayed around 45-51% for stays up to one year but dropped to 34.7% for stays longer than one year. This makes sense because hospital detox is short, outpatient patients live at home so its easier to stop coming, and treatments that last longer than a year tend to have something go wrong so fewer people finish.

Extra Credit (optional, up to 2 pts)

#### Again, this is meant to *stretch* you, not *stress* you!

Make one visualization that shows service type and length of stay **together**–that is, how completion rate varies across both at once. Faceting is one way, but there are others.

Then, in a sentence, say what this combined view shows that your two separate charts in Question 3 could not.

```{r}

#| label: extra-credit

completion_by_both <- teds_d |>
  mutate(los_group = cut(
    los,
    breaks = c(0, 1, 7, 30, 32, 34, 36, 37),
    labels = c("1 day", "2–7 days", "8–30 days", "31–60 days",
               "61–120 days", "121–365 days", "Over 1 year")
  )) |>
  filter(!is.na(los_group), services_d != "unknow") |>
  group_by(services_d, los_group) |>
  summarize(
    total     = n(),
    completed = sum(reason == "Treatment completed", na.rm = TRUE),
    .groups   = "drop"
  ) |>
  mutate(
    pct_completed = completed / total * 100,
    pct_show      = if_else(total >= 100, pct_completed, NA),   # hide tiny groups
    label         = if_else(is.na(pct_show), "", paste0(round(pct_show), "%")),
    text_col      = if_else(!is.na(pct_show) & pct_show > 55, "white", "grey15")
  )

both_p <- ggplot(completion_by_both,
                 aes(x = los_group,
                     y = fct_reorder(services_d, pct_completed, .fun = median))) +
  geom_tile_interactive(
    aes(fill = pct_show,
        tooltip = paste0(services_d,
                         "<br>Stay: ", los_group,
                         "<br>Completed: ", round(pct_completed, 1), "%",
                         "<br>Discharges: ", scales::comma(total))),
    color = "white", linewidth = 0.8) +
  geom_text(aes(label = label, color = text_col), size = 3) +
  scale_color_identity() +
  scale_fill_viridis_c(option = "mako", direction = -1, limits = c(0, 100),
                       na.value = "grey90",
                       labels = function(x) paste0(x, "%")) +
  scale_y_discrete(labels = scales::label_wrap(30)) +
  labs(title    = "Completion Rate by Service Type and Length of Stay",
       subtitle = "TEDS-D 2024 (SAMHSA)",
       caption  = "Gray tiles = fewer than 100 discharges (too few to trust)",
       x = "Length of stay",
       y = NULL,
       fill = "% completed") +
  theme_minimal(base_size = 11) +
  theme(panel.grid  = element_blank(),
        plot.title  = element_text(face = "bold"),
        axis.text.x = element_text(angle = 30, hjust = 1))

girafe(ggobj = both_p, width_svg = 9, height_svg = 5.5)
```

**What the combined view adds: T**his heatmap is a good way to look at both at once. Just by the chart, you could tell that each treatment type has its own pattern. I could tell that staying longer helps some, but others not so much. For example, staying longer helps long-term rehab (goes from 4% on the first day to 64% over 1 year), but not ambulatory detoxification (92% at 2-7 days to about 6% for staying longer)

------------------------------------------------------------------------

## Push to GitHub AND submit link on Canvas!

Assignment is due **Friday 10/2, 11:59pm**.
