---
title: "CS-201 First Look at R"
author: "Amber Camp"
date: 8/26/26
format: html
editor: visual
toc: true
---

This is the first class session walking through code together!

We will be programming in R and in this class will primarily use `qmd` documents to keep things tidy and presentable. Each code chunk below is where you will run R code. Feel free to add notes as we go along.

## Navigating RStudio

By default, everyone should have the same four-pane layout:

1.  **Top left**: Source editor

2.  **Bottom left**: Console (and Terminal and Background Jobs)

3.  **Top Right**: Environment

4.  **Bottom Right**: Files, Plots, Help, etc.

### Customizing look & feel

Feel free to customize your RStudio.\
Go to Tools \> Global Options \> Appearance to select a color scheme you like.

I also personally like to enable these:

- Tools \> Global Options \> Code \> Insert matching parens/quotes

- Tools \> Global Options \> Console \> Show syntax highlighting in console input

I don't recommend rearranging your pane layout at this time.

## Running Code

Try running some code

```{r}
library(ggplot2)
```

Some scripting

```{r}

##this is a comment 

quiz1 <- 88 #this is how to assigne a valua to a name
quiz2 <- 98
quiz3 <- 79

quiz1

#averaging
(quiz1 + quiz2 + quiz3) / 3

mean(quiz1 + quiz2 + quiz3) #not correct, just adds everything togther
mean(quiz1, quiz2, quiz3) # also wrong only returns the first arg
mean(c(quiz1, quiz2, quiz3))

list <- c(quiz1, quiz2, quiz3)

mean(list)

average <- mean(list)

quiz3 <- 99

average

list <- c(quiz1, quiz2, quiz3)
average <- mean(list)
average
round(average, 2)

```

more scores

```{r}
scores <- c(88, 94, 79, 91, 85) 
scores

mean(scores)
min(scores)
max(scores)
length(scores)

summary(scores)

students <- c("a", "b", "c", "d", "e")
students

#barplot 
barplot(scores, names.arg = students)

barplot(scores, names.arg = students, ylim = c(0,100), main = "TITLE", xlab = "pupils", y = "what they got")
```

## Introducing `ggplot2`

#### Make a dataframe

```{r}
grades_df <- data.frame(student = students, score = scores)

grades_df

summary(grades_df)

nrow(grades_df) #num of row
ncol(grades_df) #num of col
dim(grades_df) # of dimension

```

#### Try ggplot

```{r}
ggplot(data = grades_df, aes(x = student, y = score))

ggplot(data = grades_df, aes(x = student, y = score, group = 1)) +
  geom_col(fill = "steelblue") +
  geom_point() +
  geom_line() +
  theme_minimal() +
  labs(title = "BEST GRAPH EVER",
       subtitle = "some quiz scores",
       x = "da students",
       y = "what they scored on their quiz")
```

#### Mutate the dataframe

```{r}
library(tidyverse)

grades_df <- grades_df |> 
  mutate(study_hours = c(2, 5, 1, 4, 3))
```

#### Make a scatterplot

```{r}
ggplot(data = grades_df, aes(x = study_hours, y = score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, colour = "Pink")

#connot use fill (use color instead), dont use reorder
```

## Extra Credit (1pt)

What result do you expect when you run this expression?

```{r}

1 + 1 + 1

quiz1 + quiz2

1 + 1 + 1 - 3 

0.1 + 0.1 + 0.1 - 0.3

options(scipen = 999)

print(0.1, 20)
print(0.3, 20)
print(0.5, 20)
```

What did you observe?

> Answer:

Why do you think this is happening? (You can look online for help)

> Answer:
