---
title: "CS-201 First Look at R"
author: "Amber Camp"
date: 8/26/26
format: html
editor: visual
toc: true
---

This is the first class session walking through code together!

We will be programming in R and in this class will primarily use `qmd` documents to keep things tidy and presentable. Each code chunk below is where you will run R code. Feel free to add notes as we go along.

## Navigating RStudio

By default, everyone should have the same four-pane layout:

1.  **Top left**: Source editor

2.  **Bottom left**: Console (and Terminal and Background Jobs)

3.  **Top Right**: Environment

4.  **Bottom Right**: Files, Plots, Help, etc.

### Customizing look & feel

Feel free to customize your RStudio.\
Go to Tools \> Global Options \> Appearance to select a color scheme you like.

I also personally like to enable these:

- Tools \> Global Options \> Code \> Insert matching parens/quotes

- Tools \> Global Options \> Console \> Show syntax highlighting in console input

I don't recommend rearranging your pane layout at this time.

## Running Code

Try running some code

```{r}
library(ggplot2)
```

Some scripting

```{r}

##this is a comment 

quiz1 <- 88 #this is how to assigne a valua to a name
quiz2 <- 98
quiz3 <- 79

quiz1

#averaging
(quiz1 + quiz2 + quiz3) / 3

mean(quiz1 + quiz2 + quiz3) #not correct, just adds everything togther
mean(quiz1, quiz2, quiz3) # also wrong only returns the first arg
mean(c(quiz1, quiz2, quiz3))

list <- c(quiz1, quiz2, quiz3)

mean(list)

average <- mean(list)

quiz3 <- 99

average

list <- c(quiz1, quiz2, quiz3)
average <- mean(list)
average
round(average, 2)

```

more scores

```{r}
scores <- c(88, 94, 79, 91, 85) 
scores

mean(scores)
min(scores)
max(scores)
length(scores)

summary(scores)

students <- c("a", "b", "c", "d", "e")
students

#barplot 
barplot(scores, names.arg = students)

barplot(scores, names.arg = students, ylim = c(0,100), main = "TITLE", xlab = "pupils", y = "what they got")
```

## Introducing `ggplot2`

#### Make a dataframe

```{r}
grades_df <- data.frame(student = students, score = scores)

grades_df

summary(grades_df)

nrow(grades_df) #num of row
ncol(grades_df) #num of col
dim(grades_df) # of dimension

```

#### Try ggplot

```{r}
ggplot(data = grades_df, aes(x = student, y = score))

ggplot(data = grades_df, aes(x = student, y = score, group = 1)) +
  geom_col(fill = "steelblue") +
  geom_point() +
  geom_line() +
  theme_minimal() +
  labs(title = "BEST GRAPH EVER",
       subtitle = "some quiz scores",
       x = "da students",
       y = "what they scored on their quiz")
```

#### Mutate the dataframe

```{r}
library(tidyverse)

grades_df <- grades_df |> 
  mutate(study_hours = c(2, 5, 1, 4, 3))
```

#### Make a scatterplot

```{r}
ggplot(data = grades_df, aes(x = study_hours, y = score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, colour = "Pink")

#connot use fill (use color instead), dont use reorder
```

## Extra Credit (1pt)

What result do you expect when you run this expression?

```{r}

1 + 1 + 1

quiz1 + quiz2

1 + 1 + 1 - 3 

0.1 + 0.1 + 0.1 - 0.3

options(scipen = 999)

print(0.1, 20)
print(0.3, 20)
print(0.5, 20)
```

What did you observe?

> Answer: The simple arithmetic expressions like "1+1+1" and "1+1+1-3", both gave the exact results I expected. When I printed "0.1+0.1+0.1-0.3", it did equal 0, instead it gave a tiny nonzero number. When I also print 0.1, 0.3, 0.5 out to 20 decimal places, I saw how 0.1 and 0.3 are not stored as exactly those values. BUT 0.5 was stored exactly the same.

Why do you think this is happening? (You can look online for help)

> Answer: I totally remember learning something about this in my number theory class I took. Our computers only count in binary which is just ones and zeros. in binary, a fraction only ends cleanly if its denominator is a powers of 2, so 2, 4, 8, 16, 32, etc. Lets look at 0.5 (in fractions it is 1/2). Since 2 is a power of 2, this fraction converts perfectly into binary without any leftover. As for 0.1, the fraction is 1/10 and 10 is not a power of 2. It breaks down into 10 = 2 \* 5, the "5" is the problem. Instead of ending cleanly, 0.1 turns into a binary fraction that repeats forever. And this is the same for 0.3.
